A White Noise Approach to Evolutionary Ecology

### 2020-04-11 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
the models name is SAGA
expanding on coexistence

\begin{multline}
N(t)=N(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)\cdot1+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}1\right)+1\sqrt{V\nu(x,s)}\dot W(x,s)dsdx \\
=N(0)+\int_0^t\bar m(s)N(s)dt+\int_0^t\sqrt{VN(s)}d\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)}),
\end{multline}
where
\begin{equation}
\bar m(t)=\frac{1}{N(t)}\int_\mathbb{R}m(\nu,x)\nu(x,t)dx,
\end{equation}
and
\begin{equation}
\int_0^td\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)})=\int_0^t\int_\mathbb{R}\frac{\sqrt{\nu(x,s)}}{\sqrt{\int_\mathbb{R}\nu(x,s)dx}}\dot W(x,s)dxds.
\end{equation}
Setting $W_1(t)=\hat{\mathbf{W}}_t(\sqrt{\nu(x,t)})$, we can use traditional stochastic differential notation to write
\begin{equation}
dN=\bar mNdt+\sqrt{VN}dW_1.
\end{equation}
To find the associated SDE for $\bar x(t)$ and $\sigma^2(t)$, we want to repeat the same approach for $f(x)=x,x^2$ and apply It&ocirc;'s lemma. However, for these cases $f\notin C^2_b(\mathbb{R})$ since $f$ will not be bounded. However, if we can show that $\int_\mathbb{R}(|x|+x^2+x^4)\nu(x,t)dx<+\infty$ for all $t>0$ given this condition is satisfied by $\nu(x,0)$, then we can apply the mild solution of (\ref{SPDE}) to derive SDE for $\bar x(t)$ and $\sigma^2(t)$. To illustrate, let us suppose this is the case. Setting $\tilde x(t)=\int_\mathbb{R}x\nu(x,t)dx$, we have
\begin{equation}
\tilde x(t)=\tilde x(0)+\int_0^t\int_\mathbb{R}\nu(x,s)m(\nu,x)x+x\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Similarly, setting $\tilde{\tilde\sigma}^2(t)=\int_\mathbb{R}x^2\nu(x,t)dx$, we have
\begin{equation}
\tilde{\tilde\sigma}^2(t)=\tilde{\tilde\sigma}^2(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)x^2+\mu\right)+x^2\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Since $\bar x(t)=\tilde x(t)/N(t)$ and $\sigma^2(t)=\tilde{\tilde\sigma}^2(t)/N(t)-\bar x^2(t)$, we can use It&ocirc;'s lemma to derive SDE for $\bar x(t)$ and $\sigma^2(t)$, which we perform in SM \S\ref{SDE_DERIV}. We do not make any general assertions about the existence or uniqueness of $\bar x(t)$ or $\sigma^2(t)$.

@@

In SM \S\ref{SDE_DERIV} we show SDE for $N(t)$, $\bar x(t)$ and $\sigma^2(t)$ can be expressed as
\begin{subequations}\label{gen_eqns}
\begin{equation}\label{N}
dN(t)=\bar m(t)N(t)dt+\sqrt{V N(t)}dW_1(t),
\end{equation}
\begin{equation}\label{xbar_gen}
d\bar x(t)=\mathrm{Cov}_t\Big(x,m(\nu,x)\Big)dt+\sqrt{V\frac{\sigma^2(t)}{N(t)}}dW_2(t),
\end{equation}
\begin{equation}\label{sig2_gen}
d\sigma^2(t)=\mathrm{Cov}_t\Big((x-\bar x(t))^2,m(\nu,x)\Big)dt+\left(\mu-V\frac{\sigma^2(t)}{N(t)}\right)dt+\sqrt{V\frac{\overline{(x-\bar x(t))^4}-\sigma^4(t)}{N(t)}}dW_3(t),
\end{equation}
\end{subequations}
where $W_1$, $W_2$ and $W_3$ are standard Brownian motions. We note that conditions on the growth rate $m$ to guarantee existence and uniqueness of solutions to (\ref{xbar_gen}) and (\ref{sig2_gen}) have yet to be investigated. However, our results on the deterministic PDE suggest that $m(y,x)$ bounded above and differentiable in both arguments is sufficient. Dividing by $dt$ one can interpret equations (\ref{gen_eqns}) as if they are ordinary differential equations, but this not technically rigorous since Brownian motion is nowhere differentiable with respect to time. In SM \S\ref{SDE_DERIV} we show that in general $W_1$ is independent of both $W_2$ and $W_3$, but $W_2$ and $W_3$ covary.

@@

Note that although $r_i(t)$ is referred to in the coexistence literature as the intrinsic growth rate of the population, $R_i$ is a deeper intrinsic quantity. For now we refer to $R_i$ as the _innate_ growth rate. Previous work has shown the importance of demographic stochasticity [@Schreiber2017] and evolutionary dynamics [@Schreiber2018] in determining ceoxistence of competing species. However, these studies consider evolution and demographic stochasticity separately. With the connection outlined in this section formally established, researchers may pursue a postmodern coexistence theory that naturally includes both the evolutionary dynamics of populations and the effects of demographic stochasticity in a simple synthetic framework.

@@

As noted by Feller [-@feller1951], diffusion limits provide reasonable approximations for large populations, but relatively small populations require discrete models. Hence, as a diffusion limit, SPDE (\ref{SPDE}) restricts parameter values to regions that maintain large population sizes. Although this suggests an important restriction on any model developed under this framework by implying populations are not at risk of extinction, the SDE describing abundance dynamics technically permits extinction. However, for small abundances, pathology emerges in the evolution of trait means and variances. In particular, stochastic components of the SDE describing the evolution of $\bar x$ and $\sigma^2$ diverge towards infinite values as $N\to0$. Hence, studies of evolutionary rescue and colonization-extinction dynamics that incorporate phenotypic evolution should be pursued via a different approach. A natural alternative can be developed utilizing the underlying BBM that SPDE (\ref{SPDE}) is a diffusion limit of (see section \ref{stochastic}). This process explicitly tracks individuals throughout their life-history and captures reproduction as branching events. Hence, BBM processes model population size as a non-negative integer instead of a continuously varying number. In particular, the pathological behavior described for the diffusion limit does not occur for BBM as population size goes to zero. Relating this concern to our comment in \S\ref{ecoevo} on using our model of diffuse coevolution to extend modern coexistence theory [sensu @Chesson2000] in a way that accounts for phenotypic evolution and abundance dynamics, it seems necessary to determine whether or not results pertaining to the coexistence of competing populations based on deterministic models that ignore phenotypic evolution hold using stochastic individual-based simulations. To stay within the framework outlined in this paper and thereby produce comparable results, these simulations should be extensions of branching processes such as BBM that account for the effects of competition. An extension of BBM that accounts for competition and stabilizing selection was used to produce Figure \ref{rescaled} in \S\ref{stochastic}. Code for this simulation can be found in the associated github repository or by contacting the lead author of this paper.

>>>>

\begin{multline}
N(t)=N(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)\cdot1+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}1\right)+1\sqrt{V\nu(x,s)}\dot W(x,s)dsdx \\
=N(0)+\int_0^t\bar m(s)N(s)dt+\int_0^t\sqrt{VN(s)}d\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)}),
\end{multline}
where
\begin{equation}
\bar m(t)=\frac{1}{N(t)}\int_\mathbb{R}m(\nu,x)\nu(x,t)dx,
\end{equation}
and
\begin{equation}
\int_0^td\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)})=\int_0^t\int_\mathbb{R}\frac{\sqrt{\nu(x,s)}}{\sqrt{\int_\mathbb{R}\nu(x,s)dx}}\dot W(x,s)dxds.
\end{equation}
Setting $W_1(t)=\hat{\mathbf{W}}_t(\sqrt{\nu(x,t)})$, we can use traditional stochastic differential notation to write
\begin{equation}
dN=\bar mNdt+\sqrt{VN}dW_1.
\end{equation}
To find the associated SDE for $\bar x(t)$ and $\sigma^2(t)$, we want to repeat the same approach for $f(x)=x,x^2$ and apply It&ocirc;'s lemma. However, for these cases $f\notin C^2_b(\mathbb{R})$ since $f$ will not be bounded. But if we can show $\int_\mathbb{R}(|x|+x^2+x^4)\nu(x,t)dx<+\infty$ for all $t>0$ given this condition is satisfied by $\nu(x,0)$, then we can apply the mild solution of (\ref{SPDE}) to derive SDE for $\bar x(t)$ and $\sigma^2(t)$. To illustrate, let us suppose this is the case. Setting $\tilde x(t)=\int_\mathbb{R}x\nu(x,t)dx$, we have
\begin{equation}
\tilde x(t)=\tilde x(0)+\int_0^t\int_\mathbb{R}\nu(x,s)m(\nu,x)x+x\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Similarly, setting $\tilde{\tilde\sigma}^2(t)=\int_\mathbb{R}x^2\nu(x,t)dx$, we have
\begin{equation}
\tilde{\tilde\sigma}^2(t)=\tilde{\tilde\sigma}^2(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)x^2+\mu\right)+x^2\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Since $\bar x(t)=\tilde x(t)/N(t)$ and $\sigma^2(t)=\tilde{\tilde\sigma}^2(t)/N(t)-\bar x^2(t)$, we can use It&ocirc;'s lemma to derive SDE for $\bar x(t)$ and $\sigma^2(t)$, which we perform in SM \S\ref{SDE_DERIV}. We make no attempt in finding sufficient conditions for $\int_\mathbb{R}(|x|+x^2+x^4)\nu(x,t)dx<+\infty$ to hold and hence make no general assertions about the existence or uniqueness of $\bar x(t)$ or $\sigma^2(t)$. Regardless, we will later assume $\nu(x,t)$ can be approximated by a Gaussian curve in $x$ for all $t\geq0$. This assumption implies $\int_\mathbb{R}|x|^n\nu(x,t)dx<+\infty$ for all $n\in\mathbb{N}$ and for all $t\geq0$.

@@

In SM \S\ref{SDE_DERIV} we show SDE for $N(t)$, $\bar x(t)$ and $\sigma^2(t)$ can be expressed as
\begin{subequations}\label{gen_eqns}
\begin{equation}\label{N}
dN(t)=\bar m(t)N(t)dt+\sqrt{V N(t)}dW_1(t),
\end{equation}
\begin{equation}\label{xbar_gen}
d\bar x(t)=\mathrm{Cov}_t\Big(x,m(\nu,x)\Big)dt+\sqrt{V\frac{\sigma^2(t)}{N(t)}}dW_2(t),
\end{equation}
\begin{equation}\label{sig2_gen}
d\sigma^2(t)=\mathrm{Cov}_t\Big((x-\bar x(t))^2,m(\nu,x)\Big)dt+\left(\mu-V\frac{\sigma^2(t)}{N(t)}\right)dt+\sqrt{V\frac{\overline{(x-\bar x(t))^4}-\sigma^4(t)}{N(t)}}dW_3(t),
\end{equation}
\end{subequations}
where $W_1$, $W_2$ and $W_3$ are standard Brownian motions. We note that conditions on the growth rate $m$ to guarantee existence and uniqueness of solutions to (\ref{xbar_gen}) and (\ref{sig2_gen}) have yet to be investigated. However, our results on the deterministic PDE suggest that $m(h,x)$ bounded above and differentiable in both arguments is sufficient. Dividing by $dt$ one can interpret equations (\ref{gen_eqns}) as if they are ordinary differential equations, but this not technically rigorous since Brownian motion is nowhere differentiable with respect to time. In SM \S\ref{SDE_DERIV} we show that in general $W_1$ is independent of both $W_2$ and $W_3$, but $W_2$ and $W_3$ covary.

@@

Note that although $r_i(t)$ is referred to in the coexistence literature as the intrinsic growth rate of the population, $R_i$ is a deeper intrinsic quantity. For now we refer to $R_i$ as the _innate_ growth rate. Previous work, to list just a few citations in this literature, has shown the importance of demographic stochasticity [@Turelli1980; @Schreiber2017] and evolutionary dynamics [@Case1986; @Schreiber2018] in determining coexistence of competing species. However, these studies consider evolution and demographic stochasticity separately. To our knowledge, there are no rigorous theoretical investigations of coexistence that account for evolutionary dynamics of quantitative characters and demographic stochasticity simultaneously. In fact, it appears that theoretical investigations combining these processes in a single species model to understand persistence have only just begun [@Gomulkiewicz2017]. With the connection outlined in this section formally established, researchers may pursue a postmodern coexistence theory that naturally includes both the evolutionary dynamics of populations and the effects of demographic stochasticity in a simple synthetic framework.

@@

As noted by Feller [-@feller1951], diffusion limits provide reasonable approximations for large populations, but relatively small populations require discrete models. Hence, as a diffusion limit, SPDE (\ref{SPDE}) restricts parameter values to regions that maintain large population sizes. Although this suggests an important restriction on any model developed under this framework by implying populations are not at risk of extinction, the SDE describing abundance dynamics technically permits extinction. However, for small abundances, pathology emerges in the evolution of trait means and variances. In particular, stochastic components of the SDE describing the evolution of $\bar x$ and $\sigma^2$ diverge towards infinite values as $N\to0$. Hence, studies of evolutionary rescue and colonization-extinction dynamics that incorporate phenotypic evolution should be pursued via a different approach. A natural alternative can be developed utilizing the underlying BBM that SPDE (\ref{SPDE}) is a diffusion limit of (see section \ref{stochastic}). This process explicitly tracks individuals throughout their life-history and captures reproduction as branching events. Hence, BBM processes model population size as a non-negative integer instead of a continuously varying number. In particular, the pathological behavior described for the diffusion limit does not occur for BBM as population size goes to zero. This concern relates to our comment in \S\ref{ecoevo} on using our model of diffuse coevolution to extend modern coexistence theory [sensu @Chesson2000] in a way that accounts for phenotypic evolution and demographic stochasticity. Just as previous work has sought to understand the effects of demographic stochasticity and phenotypic evolution separately, it seems necessary to determine whether or not results pertaining to the coexistence of competing populations based on deterministic models that ignore phenotypic evolution hold using stochastic individual-based simulations. To stay within the framework outlined in this paper and thereby produce comparable results, these simulations should be extensions of branching processes such as BBM that account for the effects of competition. An extension of BBM that accounts for competition and stabilizing selection was used to produce Figure \ref{rescaled} in \S\ref{stochastic}. Code for this simulation can be found in the associated github repository or by contacting the lead author of this paper.
